Limited physiological plasticity and reduced relative variance in physiological rates of ectotherm populations under climate change

### 2024-07-07 ms.qmd
Re-read and revised a few sections

---
title: "Limited physiological plasticity and reduced relative variance in physiological rates of ectotherm populations under climate change"
bibliography: ./bib/refs.bib
csl: ./bib/ecology-letters.csl
format:
  docx:
    reference-doc: ./bib/template.docx
    output-file: "ms_revised.docx"
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

@@

<p style="text-align: center;">Daniel W.A. Noble^1^ $\ddagger$, Fonti Kar^2^,  Alex Bush^4^, Frank Seebacher^3^ $\dagger$, & Shinichi Nakagawa^2^ $\dagger$</p>

## Affiliations:
^1^ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2600, Australia &NewLine;  
^2^ Ecology and Evolution Research Centre, School of Biological, Earth and Environmental Sciences, University of New South Wales, Sydney, NSW, Australia &NewLine;  
^3^ School of Life and Environmental Sciences, University of Sydney, Sydney, NSW 2006, Australia  &NewLine;  
^4^ Department of Biology, Lancaster University, Liverpool, UK &NewLine;  
$\dagger$ contributed equally &NewLine;  
$\ddagger$ corresponding author, daniel.noble@anu.edu.au &NewLine;

@@

Climate change is expected to result in warmer and more variable thermal environments globally. Maintaining variability in physiological rates and adjusting them in response to extreme temperatures (plasticity) is essential for allowing populations to adapt to climate change. Yet, incorporating both plasticity and changes in phenotypic variation when predicting the impacts of climate change on populations is often ignored. We developed effect size estimates that quantify how physiological rate variance changes relative to the mean as temperature increases, which we refer to as 'relative variance'. We show by meta-analysing >1900 effects from 226 species that relative variance in physiological rates decreases at higher temperatures driven by the lower proportional increase in physiological rate variance relative to the mean. Freshwater ectotherms are capable of acclimating and have the smallest reductions in relative variance. Marine organisms also acclimated, but plasticity traded-off with a reduction in relative variance in physiological rates at higher temperatures. Relative variance reductions were particularly pronounced for terrestrial ectotherms, and this coincided with a lack of capacity for acclimation, highlighting the vulnerability of terrestrial ectotherms to climate change. Our results show that, while there are weak acclimation responses in some habitats (marine and freshwater), such responses may trade-off with reductions in relative physiological rate variance that may have important evolutionary and ecological ramifications that affect our understanding of how climate change will impact populations now and in the future.

@@

As mean physiological rates increase with temperature there is reason to believe that the intrapopulation variability might also be impacted. NEED MORE ABOUT WHY VARIANCE NOT JUST IMPORTANCE

@@

## **Materials and Methods**
### *Literature collection*
We compiled literature on ectothermic animals that measured physiological rates (e.g., metabolic rate) at two or more temperatures after having been acclimated (or acclimatized) at these temperatures. We used data from a previous meta-analysis [@Seebacher2015-wi] and updated @Seebacher2015-wi's data by extracting data from suitable studies from our own searches that followed the same search protocol. For full details on the search protocol see the *Supplementary Materials*, where we also provide a PRISMA flow diagram of our extraction process (@fig-prisma).

@@

### *$Q_{10}$ Based Effect Sizes and Sampling Variances for Means and Variances*

@@

#### *Comparing variance in physiological rates*
@Nakagawa2015-fo proposed analogous effect size estimates to *lnRR* that allow for comparisons of changes in variance between two groups, the log variance ratio (*lnVR*) and the log coefficient of variation (*lnCVR*). *lnVR* and *lnCVR* are ratios that describe the relative difference in trait variability between two groups. We refer readers to @Nakagawa2015-fo for the equations describing *lnVR* and *lnCVR*, but these can easily be extended to their $Q_{10}$ analogues (and associated sampling variance) as follows:

$$
lnVR_{Q_{10}} = ln\left( \frac{SD_{2}}{SD_{1}} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }
$${#eq-lnq10VR}
$$
s_{lnVR_{Q_{10}}} = \left( \frac{1}{2\left( N_{2}-1 \right)} + \frac{1}{2\left(N_{1} - 1\right)} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }^2
$$ {#eq-slnq10VR}

 @eq-lnq10VR and @eq-slnq10VR describe the change in physiological rate variance (@eq-lnq10VR) across a 10&deg;C temperature change along with its sampling variance (@eq-slnq10VR). While this is a useful metric, as discussed by @Nakagawa2015-fo there is often a strong mean-variance relationship that needs to be accounted for in analysing changes in variance. As such, we calculated the coefficient of variation, which standardizes changes in variance for changes in means as follows:

@@

#### *Calculating acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates*
$R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ can all be taken from samples of organisms measured acutely at two temperatures or after having been acclimated these same temperatures (@fig-concept A). For studies that measure acute and acclimated responses we used the mean, standard deviation, and sample size to derive both acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates. For studies that only measured $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ acutely we could only calculate acute versions of these effect size estimates. For all effect sizes the higher temperature was in the numerator and the lower of the two temperatures in the denominator. As such, positive effect sizes suggest that the mean (i.e., $lnRR_{Q_{10}}$) or relative variance (i.e., $lnCVR_{Q_{10}}$) is larger at the higher of the two temperatures (numerator) when standardized to 10&deg;C. Importantly, $lnRR_{Q_{10}}$, $lnCVR_{Q_{10}}$ and $lnVR_{Q_{10}}$, as with $Q_{10}$ more generally, all assume that the effect of temperature on physiological rates (or changes in relative variance) is log-linear (see *Supplementary Materials* for further discussion). We test and control for any violations of these assumptions in our analysis (see below) (@fig-concept B).

@@

We recorded or derived a series of moderator variables from each study that are expected to have an impact on our effect size estimates. This included the duration of acclimation in days given that acclimation responses are expected to depend on how long chronic temperature exposure occurs (longer exposure = better acclimation response) [@Seebacher2015-wi]. We also recorded if the sample of animals were derived from captive or wild stocks, the life-history stage of the animals used ("adult" or "juvenile") and the habitat type ("freshwater", "marine" or "terrestrial") given that @Seebacher2015-wi show that these factors can impact $Q_{10}$ estimates. Physiological rate measures varied widely across the studies but could generally be grouped into discrete trait categories [@Seebacher2015-wi]. While we explored differences across more detailed trait categories (see Supplemental Materials),  most traits could be categorised into two broad categories that included whole-organism measures (i.e., ‘cardiac’, ‘muscle’, ‘sprint’, ‘endurance’, ‘max MR’, ‘rest MR’; see Supplemental Materials for details), which all integrate a diversity of physiological and biochemical processes. The other major trait category included measures of enzymatic reaction rates (i.e., ‘metabolic enzyme’, ‘ATPase’, 'Proton Leak' and ‘OXPHOS’) targeting specific biochemical processes. We expected that whole-organism traits would be more likely to maintain relative variation in physiological function because they are composed of a greater number of biochemical reactions and would be more likely to acclimate. We did not include traits that could not be categorised into these two (i.e., ‘Other’).

@@

Monthly average temperature data were extracted from the ERA5 climate model, available from the Copernicus climate data store [@Hersbach2020-hx]. For each population and species in the dataset we extracted a 72-year period (1950-2022) of either surface air temperature (0.01&deg; resolution) for both terrestrial and freshwater taxa, or sea surface temperature for the marine taxa (at 0.25&deg; resolution) using the *ncdf4* R package [vers. `r utils::packageVersion("ncdf4")`, @Pierce2021-lc]. We chose surface temperature because we believed that it was more likely to reflect the micro-thermal environment experienced by terrestrial and freshwater ectotherms at those locations.

@@

### *Meta-Analysis*
We analysed our data using multilevel meta-analytic (MLMA) and meta-regression (MLMR) models in R (vers. `r paste0(R.Version()$major, ".", R.Version()$minor)`) using *brms* [vers. `r utils::packageVersion("brms")` @Burkner2017-wg; @Burkner2018-ur; @noauthor_2021-az] and *metafor* [vers. `r utils::packageVersion("metafor")` @Viechtbauer2010-fn].  We fit both Bayesian and frequentist approaches to ensure that our results were consistent, and to create orchard plots more easily [vers. 2.0, @Nakagawa2021-ls; @nakagawaorchard]. In addition, Bayesian methods better protect against type I errors in the presence of complex sources of non-independence [@Nakagawa2021-uf; @Song2021-ba; @Noble2017-mz]. In all cases, frequentist and Bayesian models resulted in the same conclusions. For our Bayesian models, we ran 4 MCMC chains, each with a warm-up of 1000 followed by 4000 sampling iterations keeping every 5 iterations for a minimum of 3200 samples from the posterior distribution. We used flat Gaussian priors for 'fixed' effects (i.e., $N(0, 10)$) and a student t-distribution for 'random' effects (i.e., $student_{t}(3, 0, 10)$). We checked that all MCMC chains were mixing and had converged (i.e., $R_{hat} = 1$). We compared any competing models using Akaike's Information Criteria (AIC) (if frequentist) or Wantabe Information Criteria (WAIC) (if Bayesian). We deemed models with the lowest IC value to be best supported if there was a $\Delta IC$ between the competing models of 2 or more. If two models were within 2 $\Delta IC$ units we went with the most parsimonious model.

#### *Multi-level Meta-analysis (MLMA) Models*
We first fit multi-level meta-analysis (MLMA) models (i.e., intercept-only models) for both $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$, that included study, species, trait type, and phylogeny as random effects to account for non-independence and identify sources of variability. We refer to this model structure as “Model 1” in the results. Our MLMA models allowed us to partition the variation in $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ among these key sources while accounting for total sampling variance in each. This allowed us to calculate the proportion of total heterogeneity [i.e., $I_{total}^2$; *sensu* @Nakagawa2012-oc; @Noble2022-ty] along with various $I^2$ metrics describing the proportion of variance explained by each random effect level [@Nakagawa2012-oc]. We also present 95% prediction intervals which describe the expected distribution of effects from future studies [@Nakagawa2021-ls; @Noble2022-ty].

@@

After quantifying levels of heterogeneity, we fit a series of multi-level meta-regression (MLMR) models to test our key questions. In all models, we included the same random effects as we used in our MLMA models. Acclimation time varied from `r range(data_wide$acclim_period, na.rm = TRUE)[1]` to `r range(data_wide$acclim_period, na.rm = TRUE)[2]` days (mean $\pm$ SD = `r mean(data_wide$acclim_period, na.rm = TRUE)` $\pm$ `r sd(data_wide$acclim_period, na.rm = TRUE)` days), and terrestrial ectotherms were acclimated for a much shorter duration (mean $\pm$ SD = `r acclim_time[3,2]` $\pm$ `r acclim_time[3,3]` days, n = `r acclim_time[3,4]`) than both freshwater (mean $\pm$ SD = `r acclim_time[1,2]` $\pm$ `r acclim_time[1,3]` days, n = `r acclim_time[1,4]`) and marine species (mean $\pm$ SD = `r acclim_time[2,2]` $\pm$ `r acclim_time[2,3]` days, n = `r acclim_time[2,4]`). However, rates of acclimation have been shown to be faster for many terrestrial groups compared to aquatic organisms [e.g., amphibians and reptiles have faster rates of acclimation than fishes; See @Einum2023-lz]. Nonetheless, to control for these possible differences, acclimation time was mean-centered (mean = 0) and included in all our models.

In addition to acclimation, all our models corrected for the possibility of the log-linearity assumption of our effect sizes being violated ([@fig-concept]; and see Supplementary Materials). We predicted that, if $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ were not strictly log-linear than we would see a decrease in average effect size for studies applying higher temperature treatments because these temperatures are expected to either converge on or cross the thermal optima of the performance curve causing reaction rates to decelerate or decrease beyond $T_{op}$ [@michaletz2024hotter]. The benefit of such an approach means that we could still take advantage of the excellent properties of $Q_{10}$ as an effect size while statistically correcting for the potential non-linearity that would be expected in the data at high treatment temperatures. Given that our data included a wide range of species and habitats, we also included a random slope of maximum temperature that varied across species because we, *a priori*, expected that each species will vary in their thermal performance curve and the chosen temperatures experimentally manipulated along the performance curve. To improve model interpretation, we also mean-centered the maximum temperature and included it in our models.

@@

We first tested the degree to which the difference in acute and acclimation $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ effects varied by habitat type (i.e., terrestrial, freshwater, and marine). Models included an interaction between effect type (i.e., acute or acclimation) and habitat (referred to as “Model 2”). Reduced mean $lnRR_{Q_{10_{acclimation}}}$ relative to $lnRR_{Q_{10_{acute}}}$ indicates that acclimation to thermal environments results in partial compensation of physiological rates (i.e., phenotypic plasticity), whereas no differences between $lnRR_{Q_{10_{acute}}}$ and $lnRR_{Q_{10_{acclimation}}}$ suggests organisms are not capable of physiological plasticity [@Havird2020-mj; @Seebacher2015-wi]. In contrast, a difference in $lnCVR_{Q_{10_{acclimation}}}$ relative to $lnCVR_{Q_{10_{acute}}}$ would imply that changes in between individual variation in physiological rates across 10&deg;C differ depending on whether acute or acclimation responses are measured.

Second, we tested whether acute and acclimation $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ differed between whole-organism versus biochemical traits across habitats by fitting an model with an interaction between type, habitat and trait category (referred to as "Model 3"). We also explored how acute and acclimation effects varied across more detailed trait categories, and these results can be found in the *Supplementary Materials*. Given that residual variance appeared to differ across trait categories we fit heteroscedastic residual variance models which were better supported in all cases ($\Delta AIC_{c}$; marine = 58, freshwater = 120, and terrestrial = 12).

@@

To understand the consequences of human-induced climate change on the potential to impact relative variance in physiological rates we fit a model that included an interaction between acclimation type, habitat type, latitude and longitude (referred to as "Model 6"). We assumed that any change in $lnCVR_{Q_{10}}$ across latitude and longitude could vary by habitat type (i.e., an interaction between habitat). We used non-linear tensors for latitude and longitude as any response could be complicated by local factors (e.g., altitude). Our model included random effects of species, trait, phylogeny and study. We predicted the expected change in $lnCVR_{Q_{10}}$ for each wild population in our dataset at the specific populations latitude and longitude. To do this, we first converted the predicted $lnCVR_{Q_{10}}$ to a 1&deg;C change as opposed to 10&deg;C as follows:

@@

@eq-lCVRpred turned the expected change across 10&deg;C to 1&deg;C. We then multiplied this predicted change by the change in air and sea surface temperatures at the locations of each population (and species) that is expected under high emissions scenarios in 2080.

@@

Across habitats the extent to which whole-organism versus biochemical processes acclimated varied ("Model 3"; @fig-btrait A-C; and see Supplement [@fig-S2] for more detailed trait analysis). Biochemical processes were more effective at acclimating compared to whole-organism traits for ectotherms occupying marine habitats (@fig-btrait B), whereas both whole-organism and biochemical traits acclimated in freshwater ectotherms (@fig-btrait C). Unsurprisingly, neither broad trait category showed evidence of acclimation in terrestrial ectotherms (@fig-btrait A).

Acclimation capacity did not vary consistently by life-history stage with no differences between $lnRR_{Q_{10_{acclimation}}}$ and $lnRR_{Q_{10_{acute}}}$ between adult and juveniles, except for marine habitats where adults appeared to be able to acclimate to a greater extent, although the effect size was small and in the same direction and magnitude of juveniles ("Model 4"; @fig-cvrlh A-C). When comparing between acute and acclimation responses between adults and juveniles there were also no differences across habitats (Adult-Juvenile (Acute): `r mean(acute_all_lh)`, 95% CI: `r quantile(acute_all_lh, 0.025)[1]` to `r quantile(acute_all_lh, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(acute_all_lh)`; Adult-Juvenile (Acclimation): `r mean(acclim_all_lh)`, 95% CI: `r quantile(acclim_all_lh, 0.025)[1]` to `r quantile(acclim_all_lh, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(acclim_all_lh)`).

@@

Thermal variability (i.e., $CV$) experienced by a population in the past did not explain acclimation capacity (@fig-5 A) or changes in physiological rate variance (@fig-5 B) among terrestrial, marine or freshwater populations ("Model 5").

@@

## Awknowledgements
We would like to thank Malgorzata Lagisz for helping sort out the phylogeny used in our analysis. We are also grateful to three anonymous reviewers who all provided highly constructive feedback that greatly improved the manuscript.

## Funding
DWAN is supported by an Australian Research Council (ARC) Future Fellowship (FT220100276) and ARC Discovery Project (DP210101152). SN is supported by ARC Discovery Project Grants (DP210100812 and DP230101248).

>>>>

---
title: "Limited physiological plasticity and reduced relative variance in physiological rates of ectotherm populations under climate change"
bibliography: ./bib/refs.bib
csl: ./bib/ecology-letters.csl
format:
  docx:
    reference-doc: ./bib/template.docx
    output-file: "ms_revised.docx"
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
editor: 
  markdown: 
    wrap: 72
---

@@

<p style="text-align: center;">

Daniel W.A. Noble^1^ $\ddagger$, Fonti Kar^2^, Alex Bush^4^, Frank
Seebacher^3^ $\dagger$, & Shinichi Nakagawa^2^ $\dagger$

</p>

## Affiliations:

^1^ Division of Ecology and Evolution, Research School of Biology, The
Australian National University, Canberra, ACT 2600, Australia
\
^2^ Ecology and Evolution Research Centre, School of Biological, Earth
and Environmental Sciences, University of New South Wales, Sydney, NSW,
Australia
\
^3^ School of Life and Environmental Sciences, University of Sydney,
Sydney, NSW 2006, Australia
\
^4^ Department of Biology, Lancaster University, Liverpool, UK
\
$\dagger$ contributed equally
\
$\ddagger$ corresponding author, daniel.noble\@anu.edu.au

@@

Climate change is expected to result in warmer and more variable thermal
environments globally. Maintaining variability in physiological rates
and adjusting them in response to extreme temperatures (plasticity) is
essential for allowing populations to adapt to climate change. Yet,
incorporating both plasticity and changes in phenotypic variation when
predicting the impacts of climate change on populations is often
ignored. We developed effect size estimates that quantify how
physiological rate variance changes relative to the mean as temperature
increases, which we refer to as 'relative variance'. We show by
meta-analysing \>1900 effects from 226 species that relative variance in
physiological rates decreases at higher temperatures driven by the lower
proportional increase in physiological rate variance relative to the
mean. Freshwater ectotherms are capable of acclimating and have the
smallest reductions in relative variance. Marine organisms also
acclimated, but plasticity traded-off with a reduction in relative
variance in physiological rates at higher temperatures. Relative
variance reductions were particularly pronounced for terrestrial
ectotherms, and this coincided with a lack of capacity for acclimation,
highlighting the vulnerability of terrestrial ectotherms to climate
change. Our results show that, while there are weak acclimation
responses in some habitats (marine and freshwater), such responses may
trade-off with reductions in relative physiological rate variance that
may have important evolutionary and ecological ramifications that affect
our understanding of how climate change will impact populations now and
in the future.

@@

As mean physiological rates increase with temperature there is reason to
believe that the intrapopulation variability might also be impacted.
NEED MORE ABOUT WHY VARIANCE IS EXPECTED TO CHANGE!

@@

## **Materials and Methods**

### *Literature collection*

We compiled literature on ectothermic animals that measured
physiological rates (e.g., metabolic rate) at two or more temperatures
after having been acclimated (or acclimatized) at these temperatures. We
used data from a previous meta-analysis [@Seebacher2015-wi] and updated
@Seebacher2015-wi's data by extracting data from suitable studies from
our own searches that followed the same search protocol. For full
details on the search protocol see the *Supplementary Materials*, where
we also provide a PRISMA flow diagram of our extraction process
(@fig-prisma).

@@

### $Q_{10}$ Based Effect Sizes and Sampling Variances for Means and Variances

@@

#### *Comparing variance in physiological rates*

@Nakagawa2015-fo proposed analogous effect size estimates to *lnRR* that
allow for comparisons of changes in variance between two groups, the log
variance ratio (*lnVR*) and the log coefficient of variation (*lnCVR*).
*lnVR* and *lnCVR* are ratios that describe the relative difference in
trait variability between two groups. We refer readers to
@Nakagawa2015-fo for the equations describing *lnVR* and *lnCVR*, but
these can easily be extended to their $Q_{10}$ analogues (and associated
sampling variance) as follows:

$$
lnVR_{Q_{10}} = ln\left( \frac{SD_{2}}{SD_{1}} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }
$$ {#eq-lnq10VR} $$
s_{lnVR_{Q_{10}}} = \left( \frac{1}{2\left( N_{2}-1 \right)} + \frac{1}{2\left(N_{1} - 1\right)} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }^2
$$ {#eq-slnq10VR}

@eq-lnq10VR and @eq-slnq10VR describe the change in physiological rate
variance (@eq-lnq10VR) across a 10°C temperature change along with its
sampling variance (@eq-slnq10VR). While this is a useful metric, as
discussed by @Nakagawa2015-fo there is often a strong mean-variance
relationship that needs to be accounted for in analysing changes in
variance. As such, we calculated the coefficient of variation, which
standardizes changes in variance for changes in means as follows:

@@

#### *Calculating acute and acclimation* $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates

$R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ can all be taken from
samples of organisms measured acutely at two temperatures or after
having been acclimated these same temperatures (@fig-concept A). For
studies that measure acute and acclimated responses we used the mean,
standard deviation, and sample size to derive both acute and acclimation
$lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates. For
studies that only measured $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$
acutely we could only calculate acute versions of these effect size
estimates. For all effect sizes the higher temperature was in the
numerator and the lower of the two temperatures in the denominator. As
such, positive effect sizes suggest that the mean (i.e.,
$lnRR_{Q_{10}}$) or relative variance (i.e., $lnCVR_{Q_{10}}$) is larger
at the higher of the two temperatures (numerator) when standardized to
10°C. Importantly, $lnRR_{Q_{10}}$, $lnCVR_{Q_{10}}$ and
$lnVR_{Q_{10}}$, as with $Q_{10}$ more generally, all assume that the
effect of temperature on physiological rates (or changes in relative
variance) is log-linear (see *Supplementary Materials* for further
discussion). We test and control for any violations of these assumptions
in our analysis (see below) ([@fig-concept]B).

@@

We recorded or derived a series of moderator variables from each study
that are expected to have an impact on our effect size estimates. This
included the duration of acclimation in days given that acclimation
responses are expected to depend on how long chronic temperature
exposure occurs (longer exposure = better acclimation response)
[@Seebacher2015-wi]. We also recorded if the sample of animals were
derived from captive or wild stocks, the life-history stage of the
animals used ("adult" or "juvenile") and the habitat type ("freshwater",
"marine" or "terrestrial") given that @Seebacher2015-wi show that these
factors can impact $Q_{10}$ estimates. Physiological rate measures
varied widely across the studies but could generally be grouped into
discrete trait categories [@Seebacher2015-wi]. While we explored
differences across more detailed trait categories (see *Supplemental
Materials* for details on the types of traits), most traits could be
categorised into two broad categories that included whole-organism
measures, which all integrate a diversity of physiological and
biochemical processes, and biochemical processes (e.g., enzyme reaction
rates, proton leak). Traits that could not be categorised into these two
we classified as ‘Other’.

@@

Monthly average temperature data were extracted from the ERA5 climate
model, available from the Copernicus climate data store
[@Hersbach2020-hx]. For each population and species in the dataset we
extracted a 72-year period (1950-2022) of either surface air temperature
(0.01° resolution) for both terrestrial and freshwater taxa, or sea
surface temperature for the marine taxa (at 0.25° resolution) using the
*ncdf4* R package [vers. `r utils::packageVersion("ncdf4")`,
@Pierce2021-lc]. We chose surface temperature because we believed that
it was more likely to reflect the micro-thermal environment experienced
by terrestrial and freshwater ectotherms at those locations.

@@

### *Meta-Analysis*

We analysed our data using multilevel meta-analytic (MLMA) and
meta-regression (MLMR) models in R (vers.
`r paste0(R.Version()$major, ".", R.Version()$minor)`) using *brms*
[vers. `r utils::packageVersion("brms")` @Burkner2017-wg;
@Burkner2018-ur; @noauthor_2021-az] and *metafor* [vers.
`r utils::packageVersion("metafor")` @Viechtbauer2010-fn]. We fit both
Bayesian and frequentist approaches to ensure that our results were
consistent, and to create orchard plots more easily [vers. 2.0,
@Nakagawa2021-ls; @nakagawaorchard]. In addition, Bayesian methods
better protect against type I errors in the presence of complex sources
of non-independence [@Nakagawa2021-uf; @Song2021-ba; @Noble2017-mz]. In
all cases, frequentist and Bayesian models resulted in the same
conclusions. For our Bayesian models, we ran 4 MCMC chains, each with a
warm-up of 1000 followed by 4000 sampling iterations keeping every 5
iterations for a minimum of 3200 samples from the posterior
distribution. We used flat Gaussian priors for 'fixed' effects (i.e.,
$N(0, 10)$) and a student t-distribution for 'random' effects (i.e.,
$student_{t}(3, 0, 10)$). We checked that all MCMC chains were mixing
and had converged (i.e., $R_{hat} = 1$). We compared any competing
models using Akaike's Information Criteria (AIC) (if frequentist) or
Wantabe Information Criteria (WAIC) (if Bayesian). We deemed models with
the lowest IC value to be best supported if there was a $\Delta IC$
between the competing models of 2 or more. If two models were within 2
$\Delta IC$ units we went with the most parsimonious model.

#### *Multi-level Meta-analysis (MLMA) Models*

We first fit multi-level meta-analysis (MLMA) models (i.e.,
intercept-only models) for both $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$,
that included study, species, trait type, and phylogeny as random
effects to account for non-independence and identify sources of
variability. We refer to this model structure as “Model 1” in the
results. Our MLMA models allowed us to partition the variation in
$lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ among these key sources while
accounting for total sampling variance in each. This allowed us to
calculate the proportion of total heterogeneity [i.e., $I_{total}^2$;
*sensu* @Nakagawa2012-oc; @Noble2022-ty] along with various $I^2$
metrics describing the proportion of variance explained by each random
effect level [@Nakagawa2012-oc]. We also present 95% prediction
intervals which describe the expected distribution of effects from
future studies [@Nakagawa2021-ls; @Noble2022-ty].

@@

After quantifying levels of heterogeneity, we fit a series of
multi-level meta-regression (MLMR) models to test our key questions. In
all models, we included the same random effects as we used in our MLMA
models. Acclimation time varied from
`r range(data_wide$acclim_period, na.rm = TRUE)[1]` to
`r range(data_wide$acclim_period, na.rm = TRUE)[2]` days (mean $\pm$ SD
= `r mean(data_wide$acclim_period, na.rm = TRUE)` $\pm$
`r sd(data_wide$acclim_period, na.rm = TRUE)` days), and terrestrial
ectotherms were acclimated for a much shorter duration (mean $\pm$ SD =
`r acclim_time[3,2]` $\pm$ `r acclim_time[3,3]` days, n =
`r acclim_time[3,4]`) than both freshwater (mean $\pm$ SD =
`r acclim_time[1,2]` $\pm$ `r acclim_time[1,3]` days, n =
`r acclim_time[1,4]`) and marine species (mean $\pm$ SD =
`r acclim_time[2,2]` $\pm$ `r acclim_time[2,3]` days, n =
`r acclim_time[2,4]`). However, rates of acclimation have been shown to
be faster for many terrestrial groups compared to aquatic organisms
\[e.g., amphibians and reptiles have faster rates of acclimation than
fishes; See @Einum2023-lz\]. Nonetheless, to control for these possible
differences, acclimation time was mean-centered (mean = 0) and included
in all our models, though it was not a strong predictor of effect size
variation (*Supplementary Materials*, @fig-accl_time).

In addition to acclimation, all our models corrected for the possibility
of the log-linearity assumption of our effect sizes being violated
([@fig-concept]; and see Supplementary Materials @fig-lnrrq10_temp). We
predicted that, if $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ were not
strictly log-linear than we would see a decrease in average effect size
for studies applying higher temperature treatments because these
temperatures are expected to either converge on or cross the thermal
optima of the performance curve causing reaction rates to decelerate or
decrease beyond $T_{op}$ [@michaletz2024hotter]. The benefit of such an
approach means that we could still take advantage of the excellent
properties of $Q_{10}$ as an effect size while statistically correcting
for the potential non-linearity that would be expected in the data at
high treatment temperatures. Given that our data included a wide range
of species and habitats, we also included a random slope of maximum
temperature that varied across species because we, *a priori*, expected
that each species will vary in their thermal performance curve and the
chosen temperatures experimentally manipulated along the performance
curve. To improve model interpretation, we also mean-centered the
maximum temperature and included it in our models.

@@

We first tested the degree to which the difference in acute and
acclimation $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ effects varied by
habitat type (i.e., terrestrial, freshwater, and marine). Models
included an interaction between effect type (i.e., acute or acclimation)
and habitat (referred to as “Model 2”). Reduced mean
$lnRR_{Q_{10_{acclimation}}}$ relative to $lnRR_{Q_{10_{acute}}}$
indicates that acclimation to thermal environments results in partial
compensation of physiological rates (i.e., phenotypic plasticity),
whereas no differences between $lnRR_{Q_{10_{acute}}}$ and
$lnRR_{Q_{10_{acclimation}}}$ suggests organisms are not capable of
physiological plasticity [@Havird2020-mj; @Seebacher2015-wi]. In
contrast, a difference in $lnCVR_{Q_{10_{acclimation}}}$ relative to
$lnCVR_{Q_{10_{acute}}}$ would imply that changes in between individual
variation in physiological rates across 10°C differ depending on whether
acute or acclimation responses are measured.

Second, we tested whether acute and acclimation $lnRR_{Q_{10}}$ and
$lnCVR_{Q_{10}}$ differed between whole-organism versus biochemical
traits across habitats by fitting an model with an interaction between
type, habitat and trait category (referred to as "Model 3"). We also
explored how acute and acclimation effects varied across more detailed
trait categories, and these results can be found in the *Supplementary
Materials*. Given that residual variance appeared to differ across trait
categories we fit heteroscedastic residual variance models which were
better supported in all cases ($\Delta AIC_{c}$; marine = 58, freshwater
= 120, and terrestrial = 12). We expected that whole-organism traits
would be more likely to maintain relative variation in physiological
function and be less likely to acclimate because whole-organism function
relies on a greater number of biochemical reactions each with different
thermal sensitivies [@fields2001protein; @angilletta2009thermal].

@@

To understand the consequences of human-induced climate change on the
potential to impact relative variance in physiological rates we fit a
model that included an interaction between acclimation type, habitat
type, latitude and longitude (referred to as "Model 6"). We assumed that
any change in $lnCVR_{Q_{10}}$ across latitude and longitude could vary
by habitat type (i.e., an interaction between habitat). We used
non-linear tensors for latitude and longitude as any response could be
complicated by local factors (e.g., altitude). Our model included random
effects of species, trait, phylogeny and study. We predicted the
expected change in $lnCVR_{Q_{10}}$ for each wild population in our
dataset at the specific populations latitude and longitude. To do this,
we first converted the predicted $lnCVR_{Q_{10}}$ to a 1°C change as
opposed to 10°C as follows:

@@

@eq-lCVRpred turned the expected change across 10°C to 1°C. We then
multiplied this predicted change by the change in air and sea surface
temperatures at the locations of each population (and species) that is
expected under high emissions scenarios in 2080.

@@

Across habitats the extent to which whole-organism versus biochemical
processes acclimated varied ("Model 3"; [@fig-btrait]A-C; and see
Supplement, @fig-S2) for more detailed trait analysis). Biochemical
processes were more effective at acclimating compared to whole-organism
traits for ectotherms occupying marine habitats ([@fig-btrait]B),
whereas both whole-organism and biochemical traits acclimated in
freshwater ectotherms ([@fig-btrait]C). Unsurprisingly, neither broad
trait category showed evidence of acclimation in terrestrial ectotherms
([@fig-btrait]A).

Acclimation capacity did not vary consistently by life-history stage
with no differences between $lnRR_{Q_{10_{acclimation}}}$ and
$lnRR_{Q_{10_{acute}}}$ between adult and juveniles, except for marine
habitats where adults appeared to be able to acclimate to a greater
extent, although the effect size was small and in the same direction and
magnitude of juveniles ("Model 4"; [@fig-cvrlh]A-C). When comparing
between acute and acclimation responses between adults and juveniles
there were also no differences across habitats (Adult-Juvenile (Acute):
`r mean(acute_all_lh)`, 95% CI: `r quantile(acute_all_lh, 0.025)[1]` to
`r quantile(acute_all_lh, 0.975)[1]`, $p_{MCMC}$ =
`r pmcmc(acute_all_lh)`; Adult-Juvenile (Acclimation):
`r mean(acclim_all_lh)`, 95% CI: `r quantile(acclim_all_lh, 0.025)[1]`
to `r quantile(acclim_all_lh, 0.975)[1]`, $p_{MCMC}$ =
`r pmcmc(acclim_all_lh)`).

@@

Thermal variability (i.e., $CV$) experienced by a population in the past
did not explain acclimation capacity ([@fig-5]A) or changes in
physiological rate variance ([@fig-5]B) among terrestrial, marine or
freshwater populations ("Model 5").

@@

## Acknowledgements

We would like to thank Malgorzata Lagisz for helping sort out the
phylogeny used in our analysis. We are also grateful to three anonymous
reviewers who all provided highly constructive feedback that greatly
improved the manuscript.

## Funding

DWAN is supported by an Australian Research Council (ARC) Future
Fellowship (FT220100276) and ARC Discovery Project (DP210101152). SN is
supported by ARC Discovery Project Grants (DP210100812 and DP230101248).

## Data and code availability
